Renewal and stability in populations structured by remaining years of life

### 2014-04-23 ThanoRepro/ThanoRepro.tex
included lifecycle graph, more text edits

\documentclass{article}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\usepackage{url}
\usepackage{nth}
% for the d in integrals
\newcommand{\dd}{\; \mathrm{d}}

@@

All demographic forces vary over age in known and regular ways. Information on
such forces, along with the size and age structure of a population allows the
demographer to make predictions about the future size and structure of a population. In this
paper, I explore some formal demographic consequences of a particular
redefinition of age. Instead of counting age as the time passed between birth and the present (or
some other moment), consider age as the amount of time left from the present
until death. Individuals in this case move in the same direction along
imaginary life lines, but the reference point for age is now at the end of the
life line instead of at the beginning. Of course, such information cannot be
requested of survey respondents or register data, because one's future time of
death is typically unknown. Instead we approximate this perspective using a
strategy similar to that used by \citet{miller2001increasing} or
\citet{lee2002approach}, which uses lifetables and aggregated
population counts to redistribute counts within age-classes over future death
times.

Temporal perspectives and rescalings of the lifecourse are abundant in
demography already. It is my view that these efforts have enriched the
discipline by producing new insights and questions much more than they have obscured it by
producing conundrums. Likely some demographic phenomena are best
described as a function of time since birth, and others of time until death,
while still others can be a function of both to some degree. Fertility and
reproduction are plausibly a function of both age perspectives, and the question
of how to partition such forces and events over time is a new conundrum that I
do not tackle here. The model presented here should therefore not be interpreted
as a suggestion for a better way to model or project population, but as an alternative way to tell the
story of population renewal within the confines of a philosophical bubble.
Thanatological and chronological renewal models, such as the Lotka-Euler renewal model, both approximate the movement of population stocks over time, but from opposite vantage points. Specifying a model of thanatological renewal expands the toolbox available to approach complex temporal population structure. I present a set of broad findings to help understand the implications of temporal structure.

@@

Our point of departure is a little formula that gives the probability of
dying at some age $y$ years in the future, given survival to some
earlier age, $a$. For age 0 this probability is simply the $d(a)$
column of the lifetable, standardized to sum to 1. For higher ages, we simply condition on survival:

@@

A convenient discretization of \eqref{eq:vaupel1} is to simply work with the
$d_a$ column of the lifetable (here single age):

@@

If $P_a$ is an age-classified discrete population count, we can derive $P_y$, a
remaining-years-classified population count, as follows:

@@

\FloatBarrier
\section*{The thanatological renewal model}
For the population model that follows, we are most interested in using the kind
of fertility rates shown in Figure~\ref{fig:FySpaghetti}, and we refer to a
unisex population based on female vital rates. Assuming constant vital rates,
the births for the present year are given by
\begin{align}
B(t) =& \int _0^\infty P(y,t)f^\star(y) \dd y = \int _0^\infty P(a,t)f(a) \dd a
\quad\text{,} \intertext{where $a$ indexes chronological age, $y$ indexes
thanatological age, $P(a), P(y)$ are population counts, and $f^\star(y), f(a)$ are
exact specific fertility probabilities (rates), which here are best thought of
as the fertility rates of females born to females. The thanatological integral can
be broken down back in terms of chronological age:} =& \int_{y=0}^\infty
\int_{a=0}^\infty f^\star(y)\frac{P(a,t)d(a+y)}{l(a)} \dd a \dd y \quad\text{,}
\intertext{where $d(a)$ is the continuous lifetable death distribution with
radix of 1, so $d(a)/l(0)$. We can relate the present population to past births
with $P(a,t) = B(t-a)l(a)$:}
\label{eq:ergostep1}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t-a)d(a+y)\dd a \dd y
\quad\text{.} \intertext{Eventually-- usually a bit quicker than is the case for
chronological age-- strong ergodicity will assert itself, and $B(t)$ will be related to
$B(t-a)$ according to a constant factor $e^{ra}$, where $r$ is the familiar
intrinsic rate of growth:}
\label{eq:ergostep2}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t)e^{-ra}d(a+y)\dd a \dd y
\quad\text{.}
\intertext{Divide out $B(t)$ to get back a familiar-looking renewal equation:}
\label{eq:thanoren}
1 =& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) d(a+y)e^{-ra}\dd a \dd y
\quad\text{.}
\end{align}
Now compare this to Lotka's chronological formulation
\begin{equation}
\label{eq:lotka}
1 = \int_{a=0}^\infty f(a)l(a)e^{-ra}\dd a \quad\text{,}
\end{equation}
and note that these are really quite similar, since $\int _{y=0}^\infty
d(a+y)\dd a = l(a)$. For intuition, notice that $l(a)$ is here split
up into pieces of $d(a)$, and imagine a 2D surface of these, where one axis is chronological age and the
other axis is thanatological age. For the chronological case \eqref{eq:lotka},
we multiply chronological age-specific fertility rates over one margin, and for
the thanatological case over the other margin. There is a strong parallel here
with the case of two-sex age-specific fertility rates. As in the case of
divergence between male and female single-sex models, there will always be divergence between
single-sex models under chronological versus thanatological age, unless the
starting population is stable. Even though the modeled population stocks are in
a way commensurable, the rates used here, $f(a)$ versus $f^\star(y)$ are
calculated on the basis of differently distributed denominators $E(a)$ versus $E(y)$.

@@

Once one finds $r$ from \eqref{eq:thanoren}, other familiar stable population
parameters can be calculated. For instance, we may calculate the mean
thanatological generation time, $T^y$:
\begin{equation}
\label{eq:Ty}
 T^y =  \frac{\int _{y=0}^\infty \int _{a=y}^\infty y e^{-ra} d(a) f^\star(y) \dd a
\dd y}{\int _{y=0}^\infty \int _{a=y}^\infty e^{-ra} d(a) f^\star(y) \dd a \dd
y} \quad \text{.}
\end{equation}
Literally, this is the mean remaining life expectancy of new mothers in the
stable population. The net reproduction rate, $R_0$ is related by, e.g.,
\begin{equation}
\label{eq:R0fromTy}
R_0 = e^{r T^y}
\end{equation}
The birth rate, $b$, is given by
\begin{equation}
\label{eq:eybrate}
b = \frac{1}{\int _{y=0}^\infty \int _{a=y}^\infty e^{-ra} d(a) \dd a
\quad\text{.} \dd y}
\end{equation}
The stable thanatological age structure, $c_y$, is the
proportion of the stable population with remaining years to live $y$:
\begin{equation}
\label{eq:cy}
c(y) = b \int _{a=y}^\infty e^{-ra} d(a) \dd a \quad\text{.}
\end{equation}
We already knew that when $r = 0$ and $y = a$, that $c(a)=c(y)$
\citep{vaupel2009life}, and now we can arrive at the same conclusion directly
via the chronological and thanatological renewal equations. For small non-zero
magnitude values of $r$ and survival patterns typical of humans, it is also the
case that $c(y|r)\approx c(a|-r)$. The relationship is asymptotically exact as
$r$ approaches zero. Other findings regarding stable age structure will be
presented in a future version of this paper. Other stable population quantities
may be estimated by similarly translating the various common definitions (e.g.,
in the glossary of \citet{coale1972growth}) to the present perspective. We will
focus on the main model rather than on these.

@@

Thanatological age-classes will ideally terminate at the highest value permitted
by data. For the data used here, there are 111 total age classes, which
translate to 111 total remaining-years classes (0-110+). In practice $\textbf{Y}$ becomes
a 111$\times$111 matrix, with most entries non-zero. Construction may appear
tedious for this reason. However, note that the bulk of fertility entries can
be derived as the outer (tensor) product $d_a \otimes f_y$, leaving only the 
first row and first column mortality discounting followed by the addition of the
survival superdiagonal. In most statistical programming languages constructing $\textbf{Y}$ entails only
a couple more lines of code than constructing a Leslie matrix.

>>>>

\documentclass{article}
\usepackage{amsmath}
\usepackage{caption}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{tikz}
%\usepackage[active,tightpage]{preview}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\usepackage{url}
\usepackage{nth}
% for the d in integrals
\newcommand{\dd}{\; \mathrm{d}}

@@

All demographic forces vary over age in known and regular ways. Information on
such forces, along with the size and age structure of a population allows the
demographer to make predictions about the future size and structure of a population. In this
paper, I explore some formal demographic consequences of a particular
redefinition of age. Instead of counting age as the time passed between birth and the present (or
some other moment), consider age as the amount of time left from the present
until death. Individuals in this case move in the same direction along
imaginary life lines, but the reference point for age is now at the end of the
life line instead of at the beginning. Of course, such information cannot be
requested of survey respondents,
because one's future time of death is typically unknown. It would be
possible to organize lifecourse data by remaining years of life in a
very long running linked population register, but such registers are few and far
between, and the analysis would be very retrospective in nature.
Instead we approximate this perspective using a strategy similar to that used by
\citet{miller2001increasing} or \citet{lee2002approach}, which uses lifetables
and aggregated population counts to redistribute counts within age-classes over
future death times. This allows us to make statements about the remaining
lifelines within a current stock of population, under certain assumptions.

Temporal perspectives and rescalings of the lifecourse are abundant in
demography already. It is my view that these efforts have enriched the
discipline by producing new insights and questions much more than they have obscured it by
producing conundrums. Likely some demographic phenomena are best
described as a function of time since birth, and others of time until death,
while still others can be a function of both to some degree. Fertility and
reproduction are plausibly a function of both age perspectives, and the question
of how to partition such forces and events over time is a new conundrum that I
do not tackle here. The model presented here should therefore not be interpreted
as a suggestion for a better way to model or project population, but as an alternative way to tell the
story of population renewal within the confines of a philosophical bubble.
Thanatological (counting down towards death) and chronological (counting up
from birth) renewal models, such as the Lotka-Euler renewal model, both
approximate the movement of population stocks over time, but from opposite
vantage points. Specifying a model of thanatological renewal expands the toolbox
available to approach complex temporal population structure. I present a set of
preliminary findings to help understand the implications of temporal structure.

@@

Before hashing out the thanatological renewal model, I present the
underlying data transformation --- a little formula that
gives the probability of dying at some age $y$ years in the future, given survival to some earlier age, $a$. For age 0 this probability is simply the $d(a)$ column of the lifetable, standardized to sum to 1. For higher ages, we simply condition on survival:

@@

A convenient discretization of \eqref{eq:vaupel1} is to simply work with the
$d_a$ column of the lifetable (here with single ages):

@@

If $P_a$ is an (chronological) age-classified discrete population count, we can
derive $P_y$, a thanatological age-classified population count, as follows:

@@

\FloatBarrier
\section*{The thanatological renewal model}
For the renewal model that follows, we are most interested in using the kind
of fertility rates shown in Figure~\ref{fig:FySpaghetti}, and we refer to a
unisex population based on female vital rates. In the following, I
explain describe the model in four different ways: as a continuous
model, as a matrix model, with a lifecylcle graph, and intuitively by
appealing to how population pyramids and leaves (described earlier) change over
time. Assuming constant vital rates, the births for the present year are given by
\begin{align}
B(t) =& \int _0^\infty P(y,t)f^\star(y) \dd y = \int _0^\infty P(a,t)f(a) \dd a
\quad\text{,} \intertext{where $a$ indexes chronological age, $y$ indexes
thanatological age, $P(a), P(y)$ are population counts, and $f^\star(y), f(a)$ are
exact specific fertility probabilities (rates), which here are best thought of
as the fertility rates of females born to females. The thanatological integral can
be broken down back in terms of chronological age:} =& \int_{y=0}^\infty
\int_{a=0}^\infty f^\star(y)\frac{P(a,t)d(a+y)}{l(a)} \dd a \dd y \quad\text{,}
\intertext{where $d(a)$ is the continuous lifetable death distribution with
radix of 1, so $d(a)/l(0)$. We can relate the present population to past births
with $P(a,t) = B(t-a)l(a)$:}
\label{eq:ergostep1}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t-a)d(a+y)\dd a \dd y
\quad\text{.} \intertext{Eventually-- usually a bit quicker than is the case for
chronological age-- strong ergodicity will assert itself, and $B(t)$ will be related to
$B(t-a)$ according to a constant factor $e^{ra}$, where $r$ is the familiar
intrinsic rate of growth:}
\label{eq:ergostep2}
=& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) B(t)e^{-ra}d(a+y)\dd a \dd y
\quad\text{.}
\intertext{Divide out $B(t)$ to get back a familiar-looking renewal equation:}
\label{eq:thanoren}
1 =& \int_{y=0}^\infty \int_{a=0}^\infty f^\star(y) d(a+y)e^{-ra}\dd a \dd y
\quad\text{.}
\end{align}
Now compare this to Lotka's chronological formulation
\begin{equation}
\label{eq:lotka}
1 = \int_{a=0}^\infty f(a)l(a)e^{-ra}\dd a \quad\text{,}
\end{equation}
and note that these are really quite similar, since $\int _{y=0}^\infty
d(a+y)\dd a = l(a)$. For intuition, notice that $l(a)$ is here split
up into pieces of $d(a)$, and imagine a 2D surface of these, where one axis is chronological age and the
other axis is thanatological age. For the chronological case \eqref{eq:lotka},
we multiply chronological age-specific fertility rates over one margin, and for
the thanatological case over the other margin. There is a strong parallel here
with the case of two-sex age-specific fertility rates. As in the case of
divergence between male and female single-sex models, there will always be divergence between
single-sex models under chronological versus thanatological age, unless the
starting population is stable. Even though the modeled population stocks are in
a way commensurable, the rates used here, $f(a)$ versus $f^\star(y)$ are
calculated on the basis of differently distributed denominators $E(a)$ versus $E(y)$.

@@

Once one finds $r$ from \eqref{eq:thanoren}, other familiar stable population
parameters can be calculated. For instance, we may calculate the mean
thanatological generation time, $T^y$:
\begin{equation}
\label{eq:Ty}
 T^y =  \frac{\int _{y=0}^\infty \int _{a=y}^\infty y e^{-ra} d(a) f^\star(y) \dd a
\dd y}{\int _{y=0}^\infty \int _{a=y}^\infty e^{-ra} d(a) f^\star(y) \dd a \dd
y} \quad \text{.}
\end{equation}
Literally, this is the mean of the remaining lifespans of new mothers in the
stable population. The net reproduction rate, $R_0$ is related by, e.g.,
\begin{equation}
\label{eq:R0fromTy}
R_0 = e^{r T^y}
\end{equation}
The birth rate, $b$, is given by
\begin{equation}
\label{eq:eybrate}
b = \frac{1}{\int _{y=0}^\infty \int _{a=y}^\infty e^{-ra} d(a) \dd a
\quad\text{.} \dd y}
\end{equation}
The stable thanatological age structure, $c_y$, is the
proportion of the stable population with remaining years to live $y$:
\begin{equation}
\label{eq:cy}
c(y) = b \int _{a=y}^\infty e^{-ra} d(a) \dd a \quad\text{.}
\end{equation}
We already knew that when $r = 0$ and $y = a$, that $c(a)=c(y)$
\citep{vaupel2009life}, and now we can arrive at the same conclusion directly
by equating the chronological and thanatological renewal equations. For
small non-zero magnitude values of $r$ and survival patterns typical of humans, it is also the
case that $c(y|r)\approx c(a|-r)$. The relationship is asymptotically exact as
$r$ approaches zero. Other findings regarding stable age structure will be
presented in a future version of this paper. Further stable population
quantities may be estimated by similarly translating the various common definitions (e.g.,
in the glossary of \citet{coale1972growth}) to the present perspective. We will
focus on the main model rather than on these.

@@

Figure~\ref{fig:graph} depicts the projection matrix $\textbf{Y}$ in the form of
a lifecycle graph. In this graph, each thanatological age is a node. Population
members are born into any of the thanatological age classes, and then follow
clockwise around the graph on the thick blue paths in order until finally being
absorbed into death (the node labeled ``X'', not included in the projection matrix). The light gray paths represent fertility. Thse originate from any age class that has a fertility rate greater
than zero, which for this toy matrix are all age classes except the \nth{5}
(highest). The y$^{th}$ age class is subject to the fertility rate $f^\star
(y)$, and these births distribute out to all other age classes according to the
lifetable deaths distribution, $d(a)$. Hence, the rate instensity of a fertility
path is determined by the fertility rate of the origin age, and the probability of dying at the destination age. A fraction of the fertility destined for infant mortality equal
to $lambda$ leads straight to the absorbing state of death. Fertility paths from
the \nth{4} thanatological age class have been highlighted. These correspond
to the fifth column of the example matrix $\textbf{Y}$, plus an additional path
direct to death for the fraction of infant mortality prior to December \nth{31}
in year $t$. Paths that loop back into the same age class simply indicate those
offspring that are destined to die in the same year as their mother. It is
perhaps clearer to see from the density of this graph and by recalling that
thanatological fertility distributions are relatively wide (see
Figure~\ref{fig:FySpaghetti}) that there is a high degree of mixing of
thanatological age within populations via reproduction -- a strong tendency
toward stability.

\begin{figure}
\centering
\caption{The thanatological lifecycle graph}
\label{fig:graph}
\begin{tikzpicture}[->,shorten >=1pt,auto,node distance=4.3cm,
  thick,main node/.style={circle,fill=blue!20,draw,font=\sffamily\Large\bfseries}]
  \definecolor{gray}{rgb}{0.8,0.8,0.8}
  \definecolor{lightgreen}{rgb}{0,0.8,0.8}
  \node[main node] (0) {0};
  \node[main node] (5) [above right of=0] {5};
  \node[main node] (4) [right of=5] {4};
  \node[main node] (3) [below right of=4] {3};
  \node[main node] (2) [below left of=3] {2};
  \node[main node] (1) [left of=2] {1};
  \node[main node] (X) [left of = 5] {X};

  \path[every node/.style={font=\sffamily\small},draw=gray,thin]
    (3) edge [bend left] node {} (5)
        edge [bend left] node {} (4)
        edge [bend left] node {} (2)
        edge [bend left] node {} (1)
        edge [bend left] node {} (0)  
        edge [bend left] node {} (X)
        edge [loop right] node {} (3) 
    (2) edge [bend left] node {} (5)
        edge [bend left] node {} (4)
        edge [bend left] node {} (3)
        edge [bend left] node {} (1)
        edge [bend left] node {} (0)     
        edge [loop below] node {} (2) 
        edge [bend left] node {} (X)
    (1) edge [bend left] node {} (5)
        edge [bend left] node {} (4)
        edge [bend left] node {} (3)
        edge [bend left] node {} (2)
        edge [bend left] node {} (0)    
        edge [loop below] node {} (1) 
        edge [bend left] node {} (X)
    (0) edge [bend left] node {} (5)
        edge [bend left] node {} (4)
        edge [bend left] node {} (3)
        edge [bend left] node {} (2)
        edge [bend left] node {} (1)
        edge [loop left] node {} (0) 
        edge [bend left] node {} (X);

    \path[every node/.style={font=\sffamily\small}]
    (4) edge [bend right] node {$f^\star (4)d(5)$} (5)
        edge [bend left] node {$f^\star (4)d(3)$} (3)
        edge [bend left] node {$f^\star (4)d(2)$} (2)
        edge [bend left] node {$f^\star (4)d(1)$} (1)
        edge [bend left] node {$(1-\lambda)f^\star (4)d(0)$} (0)
        edge [loop above] node {$f^\star (4)d(4)$} (4)
        edge [bend left] node {$\lambda f^\star (4)d(0)$} (X);

    \path[every node/.style={font=\sffamily\small}, ultra thick, color =
    lightgreen ] (5) edge node {} (4)
    (4) edge node {} (3)
    (3) edge node {} (2)
    (2) edge node {} (1)
    (1) edge node {} (0)
    (0) edge node {} (X);

\end{tikzpicture}
\end{figure}

Thanatological age-classes will ideally terminate at the highest value permitted
by data. For the data used here, there are 111 total age classes, which
translate to 111 total remaining-years classes (0-110+). In practice $\textbf{Y}$ becomes
a 111$\times$111 matrix, with most entries non-zero (nearly complete
connectivity, such as in Figure~\ref{fig:graph}). Construction may appear tedious for
this reason. However, note that the bulk of fertility entries can be derived as the outer (tensor) product $d_a \otimes f_y$, leaving only the first row and first column mortality discounting followed by the addition of the
survival superdiagonal. In most statistical programming languages constructing $\textbf{Y}$ entails only
a couple more lines of code than constructing a Leslie matrix.
